---
title: Testing `fitTrendCV2()` on real data
author: Aaron Lun
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document
---

```{r, echo=FALSE}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

# Introduction

This vignette is designed to test the `fitTrendCV2()` function with respect to fitting on spike-in and endogenous genes.
Evaluation will be based on how well the curve fits to the observed mass of points.

```{r}
library(scran)
library(scRNAseq)
library(scater)
```

We define a function to quickly compute the trend with and without spike.ins.

```{r}
FUN <- function(sce, has.spikes=TRUE, ...) {
     # Fitting a trend:
     if (has.spikes) {
         fit.spike <- metadata(modelGeneCV2WithSpikes(sce, spikes=1))
     }
     fit.endog <- metadata(modelGeneCV2(sce))

     # Comparing the two trend fits:
     plot(fit.endog$mean, fit.endog$cv2, xlab="Mean", ylab="CV2", ..., log="xy")
     curve(fit.endog$trend(x), add=TRUE, col="dodgerblue")
     if (has.spikes) {
         points(fit.spike$mean, fit.spike$cv2, col="red", pch=16)
         curve(fit.spike$trend(x), add=TRUE, col="purple")
     }
}
```

# Lun data

First the 416B data:

```{r}
sce <- LunSpikeInData()
QC <- perCellQCMetrics(sce)
chosen <- quickCellQC(QC, percent_subsets="altexps_ERCC_percent", nmads=3)
sce <- sce[,!chosen$discard]
FUN(sce)
```

Now the trophoblast data:

```{r}
sce <- LunSpikeInData(which="tropho")
QC <- perCellQCMetrics(sce)
chosen <- quickCellQC(QC, percent_subsets="altexps_ERCC_percent", nmads=3)
sce <- sce[,!chosen$discard]
FUN(sce)
```

# Zeisel data

Assuming quality control has already been performed:

```{r}
sce <- ZeiselBrainData()
FUN(sce)
```

# 10X data

Assuming that their cell calling was sufficient:

```{r}
library(BiocFileCache)
bfc <- BiocFileCache(ask = FALSE)
raw.path <- bfcrpath(bfc, file.path("http://cf.10xgenomics.com/samples",
    "cell-exp/2.1.0/pbmc4k/pbmc4k_filtered_gene_bc_matrices.tar.gz"))
untar(raw.path, exdir=file.path(tempdir(), "pbmc4k"))

library(DropletUtils)
fname <- file.path(tempdir(), "pbmc4k/filtered_gene_bc_matrices/GRCh38")
sce <- read10xCounts(fname, col.names=TRUE)

FUN(sce, has.spikes=FALSE)
```

# Tasic data

Removing the bare minimum to get it to work:

```{r}
sce <- TasicBrainData()
lost <- rowSums(is.na(counts(altExp(sce)))) > 0
altExp(sce) <- altExp(sce)[!lost,]

QC <- perCellQCMetrics(sce)
chosen <- quickCellQC(QC, percent_subsets="altexps_ERCC_percent", nmads=3)
sce <- sce[,!chosen$discard]
FUN(sce)
```

# Segerstolpe data

Fitting separately for each patient, as each plate was pretty different.

```{r}
sce <- SegerstolpePancreasData()
QC <- perCellQCMetrics(sce)
chosen <- quickCellQC(QC, percent_subsets="altexps_ERCC_percent", nmads=3)
sce <- sce[,!chosen$discard]

for (i in unique(sce$individual)) {
    tmp <- sce[,i==sce$individual]
    tmp <- tmp[,colSums(counts(altExp(tmp))) > 0]
    FUN(tmp, main=i)
}
```

# Session information

```{r}
sessionInfo()
```
